---
title: "The Max Step"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  out.width = "100%",
  fig.asp = 0.621,
  echo = FALSE
)
```

```{r}
#| message: false
#| warning: false

library(bggjphd)
library(tidyverse)
library(GGally)
library(cowplot)
theme_set(theme_half_open())
```


# Maximum Likelihood Estimation

## Transformed parameters

Having performed the Max step and saved the ML estimates we can easily load them by fetching the object `station_estimates`. Here we plot the distribution of transformed estimates.

$$
\begin{aligned}
\psi &= \log(\mu) \\
\tau &= \log(\frac{\sigma}{\mu}) = \log(\sigma) - \log(\mu) \\
\phi &= \mathrm{logit}(\xi, \text{ scale}=1.6) - 0.8 \\
\gamma &= \mathrm{logit}(\Delta, \text{ scale}=0.02) - 0.01.
\end{aligned}
$$

These transformations make it so that $\xi$ can take values between $-0.8$ and $0.8$, and $\Delta$ can take values between $-0.01$ and $0.01$. 

```{r}
d <- station_estimates |> 
  ungroup() |> 
  select(station, par) |> 
  unnest(par)


d |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 60) +
  facet_wrap("name", scales = "free") +
  labs(
    x = NULL,
    y = NULL,
    title = "Distributions of GEV parameters from Max step"
  )
```

## GEV parameters

We might also be interested in plotting the back-transformed parameters, i.e. $\mu$, $\sigma$, $\xi$, and $\Delta$

```{r}
d |> 
  pivot_wider() |> 
  mutate(mu = exp(psi),
         sigma = exp(tau + psi),
         xi = link_shape_inverse(phi),
         delta = link_trend_inverse(gamma)) |> 
  select(-psi, -tau, -phi, -gamma) |> 
  pivot_longer(c(-station)) |> 
  mutate(name = fct_relevel(name, "mu", "sigma", "xi", "delta")) |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 100) +
  facet_wrap("name", scales = "free") +
  labs(
    x = NULL,
    y = NULL,
    title = "Distributions of backtransformed GEV parameters from Max step"
  )
```

