---
title: "The Smooth Step: Simple spatial priors"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  out.width = "100%",
  fig.asp = 0.621,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(bggjphd)
library(tidyverse)
library(progressr)
library(future)
library(bayesplot)
library(GGally)
library(scales)
library(cowplot)
library(kableExtra)
theme_set(theme_half_open())
```

The latent parameters, $\psi$, $\tau$, $\phi$, and $\gamma$, are given intrinsic random walk spatial priors, for example

$$
\psi \sim \mathcal N(\mathbf 0, \tau_\psi \cdot Q_u).
$$

Here, $Q_u$ is defined by

$$
Q_u = R \otimes I + I \otimes R,
$$

where $I$ is the identity matrix and

$$
R = \begin{bmatrix}
1 & -1 & & & & & \\
-1 & 2 & -1 & & & & \\
& -1 & 2 & -1 & & & \\
& & \ddots & \ddots & \ddots & & \\
& & &-1 &2 &-1 & \\
& & & & -1 & 1\\
\end{bmatrix}.
$$

The results were obtained by running `ms_smooth()` in parrallel on four cores with four chains each run for 4000 samples. Half of those samples were designated as warm-up and so we have a total of 8000 samples from the posterior.

```{r, eval=FALSE}
#| eval: false
#| echo: true
n_iter <- 4000
plan(multisession, workers = 4)
with_progress({
  results <- ms_smooth(n_samp = n_iter, type = "spatial")
})
plan(sequential)
```


```{r}
if (!file.exists("results.rds")) {
  n_iter <- 4000
  plan(multisession, workers = 4)
  with_progress({
    results <- ms_smooth(n_samp = n_iter)
  })
  plan(sequential)
  write_rds(results, "results.rds")
} else {
  results <- read_rds("results.rds")
  n_iter <- results |> pull(.iteration) |> max()
}
```

# Parameters

## Hyperpriors



### Log precision scale

```{r}
results |> 
  filter(.iteration > n_iter/2) |> 
  subset_draws(
    "theta", 
  ) |> 
  summarise_draws() |> 
  kable(digits = 3) |> 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
results |> 
  filter(.iteration > n_iter/2) |>
  subset_draws("theta") |> 
  mcmc_trace()
```

```{r}
results |> 
  filter(.iteration > n_iter/2) |> 
  subset_draws("theta") |> 
  mcmc_acf_bar()
```


```{r}
results |> 
  subset_draws("theta[1]") |> 
  as_tibble() |> 
  rename(value = "theta[1]") |> 
  group_by(.chain) |> 
  mutate(accept = 1 * (value != lag(value))) |> 
  ungroup() |> 
  ggplot(aes(.iteration, accept, group = .chain)) +
  geom_smooth(method = "loess", span = 0.3, se = 0) +
  scale_x_continuous(
    expand = expansion()
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(5),
    labels = label_percent(),
    expand = expansion()
  ) +
  theme(
    plot.margin = margin(t = 5, r = 35, b = 5, l = 5)
  ) +  
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Iteration",
    y = "Acceptance probability",
    title = "Acceptance probability for theta[1]"
  )
```

### On standard deviation scale

```{r}
#| fig.asp: 0.8
results |> 
  filter(.iteration > n_iter/2) |> 
  subset_draws(
    "theta", 
  ) |> 
  mcmc_hist_by_chain(
    transformations = function(x) exp(-x/2)
  )
```


## GEV Parameters

```{r}
if (!file.exists("station_mcmc_ml.rds")) {
  
  plot_dat <- results |> 
    filter(.iteration > n_iter/2) |> 
    subset_draws(
      c("psi", "tau", "phi", "gamma")
    ) |> 
    summarise_draws()
  
  plot_dat |> 
    select(variable, mcmc_mean = mean, mcmc_median = median) |> 
    mutate(station = parse_number(variable),
           variable = str_replace(variable, "\\[.*$", "")) |> 
    inner_join(
      stations,
      by = "station"
    ) |> 
    inner_join(
      station_estimates |> 
        select(station, par) |> 
        unnest(par) |> 
        rename(ml_estimate = value, variable = name),
      by = c("station", "variable")
    ) |> 
    select(station, proj_x, proj_y, latitude, longitude, variable, ml_estimate, mcmc_mean) |> 
    write_rds(file = "station_mcmc_ml.rds")
}

plot_dat <- read_rds("station_mcmc_ml.rds") |> 
  mutate(
    variable = fct_relevel(
      factor(variable),
      "psi", "tau", "phi", "gamma"
    )
  )
```

### Comparing ML and MCMC estimates


```{r}
#| fig.asp: 1
plot_dat |> 
  pivot_longer(c(ml_estimate, mcmc_mean)) |> 
  mutate(
    variable = fct_relevel(
      factor(variable),
      "psi", "tau", "phi", "gamma"
    ),
    name = fct_recode(
      factor(name),
      "Maximum Likelihood" = "ml_estimate",
      "Posterior Mean" = "mcmc_mean"
    )
  ) |> 
  ggplot(aes(value)) +
  geom_histogram() +
  facet_wrap(vars(variable, name), ncol = 2, scales = "free_x") +
  theme(
    axis.line.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Distributions of station parameters from ML and MCMC"
  )
```


```{r}
#| fig.asp: 0.8
plot_dat |> 
  ggplot(aes(ml_estimate, mcmc_mean)) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  geom_point() +
  facet_wrap("variable", scales = "free") +
  labs(
    x = "ML Estimate (Max step)",
    y = "Posterior Mean (Smooth step)",
    title = "Comparing estimates from the Max and the Smooth steps"
  )
```

### Spatial Distributions

```{r}
proj_plot <- function(data) {
  
  title <- str_c(
    "Spatial distribution of estimates for ", unique(data$variable)
  )
  
  plot_dat <- data |> 
    pivot_longer(c(ml_estimate, mcmc_mean)) |> 
    mutate(
      name = fct_recode(
        factor(name),
        "Maximum Likelihood" = "ml_estimate",
        "Posterior Mean" = "mcmc_mean"
      )
    ) |> 
    group_by(name) |> 
    mutate(value = (value - mean(value)) / sd(value)) |> 
    ungroup()
  
  
  plot_dat |> 
    ggplot(aes(proj_x, proj_y)) +
    geom_raster(aes(fill = value)) +
    scale_fill_viridis_c() +
    facet_wrap("name", nrow = 1) +
    coord_cartesian(expand = FALSE) +
    labs(
      title = title,
      fill = NULL,
      x = "X projection",
      y = "Y projection"
    )
}
```

#### Location

##### psi

```{r}
plot_dat |> 
  filter(variable == "psi") |> 
  proj_plot()
```

##### mu

```{r}
plot_dat |> 
  filter(variable == "psi") |> 
  mutate(variable = "mu") |> 
  mutate_at(vars(ml_estimate, mcmc_mean), exp) |> 
  proj_plot()
```

#### Scale

##### tau

```{r}
plot_dat |> 
  filter(variable == "tau") |> 
  proj_plot()
```

##### sigma

```{r}
plot_dat |> 
  filter(variable %in% c("tau", "psi")) |> 
  pivot_longer(c(ml_estimate, mcmc_mean)) |> 
  pivot_wider(names_from = variable, values_from = value) |> 
  mutate(sigma = exp(tau + psi)) |> 
  select(-psi, -tau) |> 
  pivot_longer(c(sigma), names_to = "variable", values_to = "value") |> 
  pivot_wider() |> 
  proj_plot()
```

#### Shape

##### phi

```{r}
plot_dat |> 
  filter(variable == "phi") |> 
  proj_plot()
```

##### xi

```{r}
plot_dat |> 
  filter(variable == "phi") |> 
  mutate(variable = "xi") |> 
  mutate_at(vars(ml_estimate, mcmc_mean), link_shape_inverse) |> 
  proj_plot()
```

#### Trend

##### gamma

```{r}
plot_dat |> 
  filter(variable == "gamma") |> 
  proj_plot()
```

##### Delta

```{r}
plot_dat |> 
  filter(variable == "gamma") |> 
  mutate(variable = "delta") |> 
  mutate_at(vars(ml_estimate, mcmc_mean), link_trend_inverse) |> 
  proj_plot()
```
